### 2020-07-15 ISSTA/concepts.tex
down to size

Horizontal determinism can be best understood by thinking of
nondeterminism detection as an unusual kind of \emph{differential
  testing} \cite{Differential,ICSEDiff}.  In differential testing, a
system is compared against a reference in order to ensure that
it behaves equivalently, at some level of abstraction, to another
implementation.  Differential testing is
extremely powerful, in that any (properly defined) divergence of
behavior indicates a functional correctness fault in (at
least) one of the systems under test, and  is widely
used for systems software such as compilers
\cite{Differential,csmith} and file systems \cite{CFV08,AMAI}.  The major limitation of differential
testing is that multiple implementations of a system are
almost as rare as good correctness specifications.
For the special case of detecting nondeterminism, however, \emph{a
  system can serve as its own reference implementation.}    The
problem, then, becomes one of deciding at what granularity the
reference equivalence will be checked:  e.g., 
processor-instruction and memory-layout determinism is seldom
necessary or even desired.  We propose two approaches to ``aligning'' an
execution with itself.

@@

However, for the specialized case of  failure determinism, no
specification is required.
Failure determinism is the following restriction on an API: {\bf If a call/action \emph{fails}, and indicates this to the caller, it should not modify system
  state in any way; changes should be ``rolled back.''}   Some behaviors of the Mac OS High Sierra
root exploit (CVE-2017-13872 \cite{applebug0}) exhibited failure
nondeterminism.  Attempting to login with the root account with an
empty password appeared to fail, then, on a repeated try, succeeded
\cite{applebug1,applebug2}.   Many library APIs are largely failure deterministic.  For instance, if
we exclude actual I/O errors, most POSIX file system calls either
succeed or do nothing; interfaces that may only partially succeed,
such as {\tt read} and {\tt write} tend to explicitly return a degree of success, rather than signalling total failure, when appropriate.

In languages with a clear mechanism for expressing failure of a call
(e.g., exceptions in Python and Java, or Option/Result types in Rust,
Haskell, and ML), failure determinism can be automatically checked.
The overhead should be even lower than for other vertical determinism, in that we can expect most failing
operations to be fast.  Checking equivalence of full
observable state, however, is still expensive, and requires defining
the observable components of a state.  A more lightweight, but (as we
will show) still powerful way of checking for failure nondeteterminism
is to simply
\emph{repeat the failing operation, and check if it still fails}.  This more restrictive notion still captures such issues as the Apple login bug.

@@

Some sources of nondeterminism unfortunately require executing a test in a \emph{new
process environment}, because the source is inherently tied to the
process in which code runs.  In terms of our formalism, an
action needs to be introduced that indicates the generation of a fresh
process.  Address Space Layout Randomization (ASLR) \cite{ASLR}  is probably the most
important source of nondeterminism that arises (only) from change in
process.  ASLR scrambles the layout of memory of the process in which
an executable runs in order to make it harder to exploit memory-safety
vulnerabilities in code, but also makes tests more flaky
%as nondeterminism.

@@

Delta-debugging \cite{DD}  is a widely used method for reducing the
size of failing tests, making them easier to understand and debug.
The core idea of delta-debugging is to take a test with some
property (usually ``it fails'')
and produce a smaller test that has the same property.
Delta-debugging as originally proposed uses a
modified binary search, but we rely only on the general structure: \cite{CReduce,onetest}:  Given a
test case, $T_{\mathit{INIT}} : \mathit{test}$ and a predicate
$\mathit{PRED}: \mathit{test} \rightarrow \mathit{bool}$, we
reduce $T_{\mathit{INIT}}$ with respect to $\mathit{PRED}$
as follows:

@@

Delta-debugging in the context of nondeterminism has two
purposes.  One is simply the usual goal of reducing the size of a
test.  Identifying the cause of nondeterminism may be very easy in a
test consisting of ten library function calls but very difficult with
more than a hundred calls.  In horizontal nondeterminism detection, however,
delta-debugging also tends to \emph{change the probability of
  nondeterministic behavior}; this can be both harmful and
beneficial.  This behavior is part of a more general (and, to our
knowledge, not previously investigated) issue: reducing a test with
respect to an arbitrary predicate $\mathit{PRED}$ (including failure,
but also code coverage, etc. \cite{icst2014,stvrcausereduce}) that \emph{only holds with a
certain probability} (the predicate itself is nondeterministic).
\begin{comment}Note
that while most descriptions of delta-debugging discuss reducing a
test with respect to \emph{test failure} (hence the term
``delta-\emph{debugging}'') we adopt the wider view of delta-debugging
as reducing the length of a test while maintaining that an arbitrary
predicate still holds.   E.g., predicates, related to code-coverage can be very useful
\cite{icst2014,stvrcausereduce}.
\end{comment}
%We want to consider
%what happens when $\mathit{PRED}$ is not deterministic.
``The test behaves flakily'' is an obvious relevant example of such a $\mathit{PRED}$.

>>>>

Horizontal determinism can be best understood by thinking of
nondeterminism detection as an unusual kind of \emph{differential
  testing} \cite{Differential}.  In differential testing, a
system is compared against a reference in order to ensure that
it behaves equivalently, at some level of abstraction, to another
implementation.  Differential testing is
extremely powerful, in that any (properly defined) divergence of
behavior indicates a functional correctness fault in (at
least) one of the systems under test, and  is widely
used for systems software such as compilers
\cite{Differential,csmith} and file systems \cite{AMAI}.  The major limitation of differential
testing is that multiple implementations of a system are
almost as rare as good correctness specifications.
For the special case of detecting nondeterminism, however, \emph{a
  system can serve as its own reference implementation.}    The
problem, then, becomes one of deciding at what granularity the
reference equivalence will be checked:  e.g., 
processor-instruction and memory-layout determinism is seldom
necessary or even desired.  We propose two approaches to ``aligning'' an
execution with itself.

@@

However, for the specialized case of  failure determinism, no
specification is required.
Failure determinism is the following restriction on an API: {\bf If a call/action \emph{fails}, and indicates this to the caller, it should not modify system
  state in any way; changes should be ``rolled back.''}   Some behaviors of the Mac OS High Sierra
root exploit (CVE-2017-13872 \cite{applebug0}) exhibited failure
nondeterminism.  Attempting to login with the root account with an
empty password appeared to fail, then, on a repeated try, succeeded.   Many library APIs are largely failure deterministic.  For instance, if
we exclude actual I/O errors, most POSIX file system calls either
succeed or do nothing; interfaces that may only partially succeed,
such as {\tt read} and {\tt write} tend to explicitly return a degree of success, rather than signalling total failure, when appropriate.

In languages with a clear mechanism for expressing failure of a call
(e.g., exceptions in Python and Java, or Option/Result types in Rust,
Haskell, and ML), failure determinism can be automatically checked.
The overhead should be even lower than for other vertical determinism, in that most failing
operations are fast.  Checking equivalence of full
observable state, however, is still expensive, and requires defining
the observable components of a state, suggesting the shortcut of simply
\emph{repeating the failing operation, and checking if it still fails}
which still catches such issues as the Apple login bug.

@@

Some sources of nondeterminism unfortunately require executing a test in a \emph{new
process environment}, because the source is inherently tied to the
process in which code runs.  In terms of our formalism, an
action needs to be introduced that indicates the generation of a fresh
process.  Address Space Layout Randomization (ASLR), which scrambles the layout of memory of the process in which
an executable runs in order to make it harder to exploit memory-safety
vulnerabilities, is probably the most
important source of nondeterminism that arises (only) from change in
process.  
%as nondeterminism.

@@

Delta-debugging \cite{DD}  is a widely used method for reducing the
size of failing tests, making them easier to understand and debug.
The core idea of delta-debugging is to take a test with some
property (usually ``it fails'')
and produce a smaller test that has the same property.
Delta-debugging as originally proposed uses a
modified binary search, but we rely only on the general structure \cite{onetest}:  Given a
test case, $T_{\mathit{INIT}} : \mathit{test}$ and a predicate
$\mathit{PRED}: \mathit{test} \rightarrow \mathit{bool}$, we
reduce $T_{\mathit{INIT}}$ with respect to $\mathit{PRED}$
as follows:

@@

Delta-debugging in the context of nondeterminism has two
purposes.  One is simply the usual goal of reducing the size of a
test.  Identifying the cause of nondeterminism may be very easy in a
test consisting of ten library function calls but very difficult with
more than a hundred calls.  In horizontal nondeterminism detection, however,
delta-debugging also tends to \emph{change the probability of
  nondeterministic behavior}; this can be both harmful and
beneficial.  This behavior is part of a more general (and, to our
knowledge, not previously investigated) issue: reducing a test with
respect to an arbitrary predicate $\mathit{PRED}$ (including failure,
but also code coverage, etc. \cite{stvrcausereduce}) that \emph{only holds with a
certain probability} (the predicate itself is nondeterministic).
\begin{comment}Note
that while most descriptions of delta-debugging discuss reducing a
test with respect to \emph{test failure} (hence the term
``delta-\emph{debugging}'') we adopt the wider view of delta-debugging
as reducing the length of a test while maintaining that an arbitrary
predicate still holds.   E.g., predicates, related to code-coverage can be very useful
\cite{stvrcausereduce}.
\end{comment}
%We want to consider
%what happens when $\mathit{PRED}$ is not deterministic.
``The test behaves flakily'' is an obvious relevant example of such a $\mathit{PRED}$.

### 2020-07-15 ISSTA/conclusions.tex
start QRS shrink
down to size

Unexpected nondeterminism of software systems frustrates users,
whether they be humans or (more importantly) other software systems.
Nondeterminism is even more pernicious in software testing,
frustrating debugging efforts (Heisenbugs \cite{Heisenbug} and
Mandelbugs \cite{GrottkeBugs,FaultTriggers} are widely loathed), and
leading to the costly problem of flaky tests
\cite{miccoflaky,listfieldtestanalysis}.
This paper proposes a formulation of types of nondeterminism, and a
practical approach to using automated test generation to detect and understand
nondeterminism, especially in the library code that underlies most
systems. In addition to traditional \emph{horizontal} nondeterminism, we introduce
the concept of \emph{vertical} nondeterminism bugs, and show their
additional specification power compared to a
strong differential testing harness.  
\begin{comment}
We implemented our approach in the
TSTL automated test generation system for Python, and demonstrated the
simplicity and utility of the approach on real-world examples.
\end{comment}

As future work, we would like to support the automatic detection of opaque
values; if a value (e.g., a timestamp) is different in \emph{every}
trace, it is likely opaque.  This would make
testing libraries with intended nondeterminism much easier. Similarly, for
vertical nondeterminism, we would like to
automatically identify idempotent operations.
%We would also like to lower the cost
%of checking process nondeterminism, perhaps by using methods taken in
%AFL \cite{aflfuzz} to avoid high startup costs for test executions.
Finally, we are interested in using test decomposition
\cite{Composition} to isolate and understand
nondeterminism in unit tests, and to mitigate flaky tests.
Reliable detection of nondeterminism is a critical tool in making such
an approach efficient and reliable.

>>>>

Unexpected nondeterminism of software systems frustrates users,
whether they be humans or (more importantly) other software systems.
Nondeterminism is even more pernicious in software testing,
frustrating debugging efforts, and
leading to the costly problem of flaky tests
\cite{miccoflaky,listfieldtestanalysis}.
This paper proposes a formulation of types of nondeterminism, and a
practical approach to using automated test generation to detect and understand
nondeterminism in Python library code.

\begin{comment}
In addition to traditional \emph{horizontal} nondeterminism, we introduce
the concept of \emph{vertical} nondeterminism bugs, and show their
additional specification power compared to a
strong differential testing harness.

We implemented our approach in the
TSTL automated test generation system for Python, and demonstrated the
simplicity and utility of the approach on real-world examples.

As future work, we would like to support the automatic detection of opaque
values; if a value (e.g., a timestamp) is different in \emph{every}
trace, it is likely opaque.  This would make
testing libraries with intended nondeterminism much easier. Similarly, for
vertical nondeterminism, we would like to
automatically identify idempotent operations.
%We would also like to lower the cost
%of checking process nondeterminism, perhaps by using methods taken in
%AFL \cite{aflfuzz} to avoid high startup costs for test executions.
Finally, we are interested in using test decomposition
\cite{Composition} to isolate and understand
nondeterminism in unit tests, and to mitigate flaky tests.
Reliable detection of nondeterminism is a critical tool in making such
an approach efficient and reliable.
\end{comment}

### 2020-07-15 ISSTA/experiments.tex
down to size

The {\tt redis-py} \cite{redispy} module implements a (very) widely-used Python interface
to the popular Redis in-memory database, cache, and
message broker \cite{redis}.  
Using TSTL's harness for {\tt redis-py}, both {\tt redis-py} and Redis
itself can be tested; unfortunately, generating stand-alone
high-coverage regression tests for them
has proven difficult, as numerous Redis commands introduce nondeterministic
behavior:  thus the resulting tests are very often \emph{flaky}.
Using AFL, we can see that {\tt redis-py} has a 
stability of only 56.26\%, a clear indicator of significant nondeterminism. Some
of the problematic commands are obvious simply by inspection (e.g.,
{\tt randomkey, srandmember}). It is also clear that 
commands producing data with a limited lifetime (e.g., {\tt expire,
  pexpire}) introduce timing-based nondeterminism.
However, that a command such as {\tt restore} takes an expiration
argument is not obvious to a non-expert, and it is very
difficult to guess whether the {\tt pipe} mechanism, which allows a
large sequence of commands to be queued up and executed
at once, introduces nondeterminism.
\begin{comment}
  Deep experience with Redis would make these
issues clear, but tests using a library are often written by those not
intimately familiar with the semantics of every library call (otherwise they would not have
been as likely to introduce flaky tests in the first place).  This is
even more true in the case of automated test generation, where the
test engineer is often chosen for expertise in test generation tools,
not in the domain of the SUT, and is seldom the
original developer.
\end{comment}

Using the original TSTL harness, just over 20\% of all {\tt redis-py}
regression tests (of length 200) generated were flaky.  We determined
that the mean time to find all (known to us, over many experimental runs)
sources of nondeterminism and produce minimal tests is 3.2 
hours.  Our answer to {\bf RQ1} is thus a strong affirmative.

@@

We removed 11 calls from the {\tt redis-py} harness, and  modified 2
calls, after
identification of sources of nondeterminism.  After making these changes, no flakiness was observed in a sample of over 2,000 length
200 tests.   Resulting code coverage loss was minimal.
Mean branch coverage  was reduced
by less than 5 branches (a less than 1\% decrease).  AFL stability was 56.26\% with the harness allowing 
nondeterminism, but rose to 98.32\% using the harness with 
nondeterministic operations removed.

The overhead for
determinism checks, with no delay between operations, is only
20\%, much lower than the expected cost of running each test twice,
answering {\bf RQ2} also in the affirmative.

As to {\bf RQ3}, Figures \ref{fig:r1}-\ref{fig:r3} show delta-debugging of a typical
{\tt redis-py} nondeterministic test, originally with 287 steps.  The initial test behaves
nondeterministically about half the time.  Unmodified delta-debugging
eventually produces a test of length 16 that only behaves
nondeterministically 24\% of the time.  The reduction takes only 38
seconds.  For the purpose of identifying commands leading to
nondeterminism, this is acceptable.  However, if we were actually
debugging a complex nondeterminism bug in Redis itself, we might want
a more reliably nondeterministic test.  Using the same parameters as
in Section \ref{sec:pubbias}, we see the same pattern.  Simply making
a predicate that ``forces'' the probability to remain high, with a
large number of samples, does not work (Figure \ref{fig:r3}) and
requires over 2,000 seconds to produce a test with an even worse
probability of nondeterminism.  Using 10 samples and 10 replications,
on the other hand, actually improves the probability of
nondeterministic behavior, and (in this case) even produces a smaller test (only 14 steps) in just
over 10 minutes.

@@

The {\tt datarray} module \cite{datarray} is a prototype
implementation for numpy arrays with named axes to improve data
management, developed by the Berkeley Institute for Data Science.  As part of its code, it provides a set of algorithms for
inference in
Bayesian belief networks \cite{russell2016artificial}.  An earlier
version of these algorithms produced nondeterministic (and in some
cases incorrect) results due to dependence on the order of values in
an iterator over a Python dictionary, on Python versions above 3.2,
until 3.6 (see Section \ref{sec:pnondet}).

\begin{comment}
Figure \ref{hashbug} shows TSTL code for generating inputs to the {\tt
  datarray} algorithms.  The {\tt flat\_sort} function is
needed because we care about actual differences in probability values,
not simply the order of list, tuple, or dictionary items.
\end{comment}
Running our TSTL
harness to test {\tt datarray} consistently requires less than 10 seconds to produce a test exhibiting process-level nondeterminism in
the {\tt calc\_marginals\_sumproduct} function (the only broken
algorithm), answering {\bf RQ1} again in the affirmative.  Reducing this 60 step test to a minimal test of only 6 steps,
showing an extremely simple input producing the issue, 
required another 92 seconds ({\bf RQ3}) on average.  Interestingly, the way that {\tt
  python-afl} is used in TSTL means that AFL stability is 100\% even
for the original, broken, code,
because the {\tt PYTHONHASHSEED} has already been chosen before the
fork in AFL.  Removing the nondeterministic call, we can see that the cost of
checking for process nondeterminism, with no delay between operations,
is high, a mean 93\% slowdown ({\bf RQ2}.  This is due to the high cost of subprocess
creation and communication. However, even with this essentially worst
case behavior (where test runtime is very low compared to process overheads),
we don't quite halve test executions.

@@

The {\tt pyfakefs} \cite{pyfakefs} module implements a fake file
system that mocks the Python file system modules, to allow Python
tests both to run faster by using an in-memory file system and to make
file system changes that would not be safe or easily performed using
real persistent storage.  Originally developed in 2006 at Google by
Mike Bland, {\tt pyfakefs} is now used in over 2,000 Python tests,
inside and outside Google \cite{pyfakefs}.
The TSTL harness for {\tt pyfakefs} has been used to detect (and
correct) over 80 faults.
%\url{https://github.com/jmcgeheeiv/pyfakefs/issues?q=label\%3ATSTL}.

\subsubsection{Manually Inserted Fault}

@@

\subsubsection{Mutation Analysis}

We wanted to determine whether there were also simple faults that
could be detected by failure nondeterminism, but \emph{not} detected
by differential testing, and quantify the additional specification
strength provided by failure determinism checks, to further answer
{\bf RQ1} in the special case of failure determinism.  We therefore used
{\tt universalmutator} \cite{RegExpMut} to produce 2,350 mutants of
the {\tt pyfakefs} core file system code.   %We restricted the generation
%to only mutate code covered during a 60 second run of the test
%harness.
\begin{comment}
  We could have applied mutation
testing to {\tt redis-py}, but we are dubious of the value of such an
analysis without a strong differential harness to compare to; the {\tt
  redis-py} testing is limited in strength, and largely useful for
producing high-coverage operation sequences from which to construct
manual unit tests.

>>>>

The {\tt redis-py} \cite{redispy} module implements a (very) widely-used Python interface
to the popular Redis in-memory database, cache, and
message broker \cite{redis}.  
Using TSTL's harness for {\tt redis-py}, both {\tt redis-py} and Redis
itself can be tested; unfortunately, generating stand-alone
high-coverage regression tests for them
has proven difficult, as numerous Redis commands introduce nondeterministic
behavior:  thus the resulting tests are very often \emph{flaky}.
Using AFL, we can see that {\tt redis-py} has a 
stability of only 56.26\%, a clear indicator of significant nondeterminism. Some
of the problematic commands are obvious simply by inspection (e.g.,
{\tt randomkey, srandmember, expire}). 
However, issues such as whether the mechanism that allows a
sequence of commands to be queued up and executed
at once introduces nondeterminism are much more subtle.
\begin{comment}
  Deep experience with Redis would make these
issues clear, but tests using a library are often written by those not
intimately familiar with the semantics of every library call (otherwise they would not have
been as likely to introduce flaky tests in the first place).  This is
even more true in the case of automated test generation, where the
test engineer is often chosen for expertise in test generation tools,
not in the domain of the SUT, and is seldom the
original developer.
\end{comment}

Using the original TSTL harness, just over 20\% of all {\tt redis-py}
regression tests (of length 200) generated were flaky.  We determined
that the mean time to find all (known to us, over many experimental runs)
sources of nondeterminism and produce minimal tests is 3.2 
hours ( {\bf RQ1}).

@@

We removed 11 calls from the {\tt redis-py} harness, and  modified 2
calls, after
identification of sources of nondeterminism.  After making these changes, no flakiness was observed in a sample of over 2,000 length
200 tests.   Resulting code coverage loss was minimal.
Mean branch coverage  was reduced
by less than 5 branches (a less than 1\% decrease).  AFL stability was 56.26\% with the harness allowing 
nondeterminism, but rose to 98.32\% using the harness with 
nondeterministic operations removed.  The overhead for
determinism checks, with no delay between operations, is only
20\%, much lower than the expected cost of running each test twice,
answering {\bf RQ2} also in the affirmative.

As to {\bf RQ3}, Figures \ref{fig:r1}-\ref{fig:r3} show delta-debugging of a typical
{\tt redis-py} nondeterministic test, originally with 287 steps.  The initial test behaves
nondeterministically about half the time.  Unmodified delta-debugging
eventually produces a test of length 16 that only behaves
nondeterministically 24\% of the time.  The reduction takes only 38
seconds.  For the purpose of identifying commands leading to
nondeterminism, this is acceptable.  However, if we were actually
debugging a complex nondeterminism bug in Redis itself, we might want
a more reliably nondeterministic test.  Using the same parameters as
in Section \ref{sec:pubbias}, we see the same pattern.  Simply making
a predicate that ``forces'' the probability to remain high, with a
large number of samples, does not work (Figure \ref{fig:r3}) and
requires over 2,000 seconds to produce a test with an even worse
probability of nondeterminism.  Using 10 samples and 10 replications,
on the other hand, improves the probability of
nondeterministic behavior, and produces a test with only 14 steps in just
over 10 minutes.

@@

The {\tt datarray} module \cite{datarray} is a prototype
implementation for numpy arrays with named axes, which also provides a set of algorithms for
inference in
Bayesian belief networks.  An earlier
version of these algorithms sometimes produced incorrect results due to dependence on the order of values in
an iterator over a Python dictionary (see Section \ref{sec:pnondet}).  
Running our TSTL
harness to test {\tt datarray} consistently requires less than 10 seconds to find process-level nondeterminism in
the {\tt calc\_marginals\_sumproduct} function, answering {\bf RQ1}
again in the affirmative.  Reducing this 60 step test to a 6-step test 
required another 92 seconds ({\bf RQ3}) on average.  The cost of
checking for process nondeterminism 
is a mean 93\% slowdown ({\bf RQ2}).

@@

The {\tt pyfakefs} \cite{pyfakefs} module implements a fake file
system that mocks the Python file system modules, to allow Python
tests both to run faster by using an in-memory file system and to make
file system changes that would not be safe or easily performed using
real persistent storage. It is used in over 2,000 Python tests
\cite{pyfakefs}.
The TSTL harness for {\tt pyfakefs} has been used to detect (and
correct) over 80 faults.
%\url{https://github.com/jmcgeheeiv/pyfakefs/issues?q=label\%3ATSTL}.

%\subsubsection{Manually Inserted Fault}

@@

%\subsubsection{Mutation Analysis}

We wanted to determine whether there were also simple faults that
could be detected by failure nondeterminism, but \emph{not} detected
by differential testing, and quantify the additional specification
strength provided by failure determinism checks, to further answer
{\bf RQ1} in the special case of failure determinism.  We therefore used
{\tt universalmutator} \cite{RegExpMut} to produce 2,350 mutants of
the core file system code.   %We restricted the generation
%to only mutate code covered during a 60 second run of the test
%harness.
\begin{comment}
  We could have applied mutation
testing to {\tt redis-py}, but we are dubious of the value of such an
analysis without a strong differential harness to compare to; the {\tt
  redis-py} testing is limited in strength, and largely useful for
producing high-coverage operation sequences from which to construct
manual unit tests.

### 2020-07-15 ISSTA/introduction.tex
down to size

Determinism is particularly important for testing and debugging, where
being able to exactly reproduce system behavior is essential to
productivity.  Developers using a test exhibiting nondeterminism
to debug a system
face a serious challenge in reasoning about causality; this
pernicious phenomenon has been sometimes referred to as the dreaded
``Heisenbug'' \cite{Heisenbug}, though the more accurate usage is
``Mandelbug'' \cite{GrottkeBugs,FaultTriggers}, since Heisenbugs, properly speaking, are bugs that disappear or
alter behavior under instrumentation.
Even more significantly, regression testing
effectiveness can be significantly reduced if tests cover code or,
worse yet, fail only
intermittently and unpredictably, as a result of environmental
factors rather than bugs in changed code.  Such behavior is
unfortunately all too common:  Gao et
al. \cite{Gao:2015:MSU:2818754.2818764} observed coverage differences
of up to 184 lines of code for the same test, and false positive rates
as high as 96\%.  Nondeterminism is often problematic for
developers, who want to  assume that library code behaves in a  predictable fashion; nondeterminism is
frequently vexing in debugging efforts; nondeterminism is \emph{often disastrous} for large-scale highly
automated testing.

@@

Complex
modern software systems usually include a large set of
\emph{regression tests}.  A regression test suite is a
set of tests that can be run against softwarre every time it
is modified, to ensure that the modification has not broken the system.  \emph{Flaky tests} \cite{miccoflaky} are regression
tests that fail in an intermittent, unreliable fashion.  The essence of a flaky
test is that, for the same snapshot of test code and code under test
(CUT), it sometimes fails and sometimes passes: the pass/fail result (disposition) of the
test is not a deterministic property of the test code, code under
test, and testing environment.  This produces three serious
problems: first, a flaky test often wastes developer time and delays
software changes by forcing the investigation of code-under-test that
is not actually incorrect.  Second, failures in flaky tests are often ignored, and therefore serious software faults may
be missed.  Finally, to mitigate these problems, flaky tests are often
run multiple times, wasting computing resources and
delaying code changes.  Flaky tests are, for us, simply a special case of general test 
nondeterminism, in particular the 
\emph{horizontal} nondeterminism 
described below.  Our focus is on detecting sources of 
flaky-ness, \emph{before} they propagate.
\begin{comment}
  As an analogy, we note that  a canary in a coal mine is of
little use if canaries frequently become ill for reasons unrelated to
the presence of toxic gases.  Mining may stop for no good reason, or
miners may learn to ignore the canary, leading to tragedy; a third,
more ``practical'' option is that miners may carry so many redundant
canaries into the coal mine that canary-care itself becomes a serious burden.
\end{comment}

>>>>

Determinism is particularly important for testing and debugging, where
being able to exactly reproduce system behavior is essential to
productivity.  Developers using a test exhibiting nondeterminism
to debug a system
face a serious challenge.
Regression testing
effectiveness can be significantly reduced if tests cover code or,
worse yet, fail only
intermittently and unpredictably.  Such behavior is
unfortunately all too common:  Gao et
al. \cite{Gao:2015:MSU:2818754.2818764} observed coverage differences
of up to 184 LOC for the same test, and false positive rates
as high as 96\%.  Nondeterminism is problematic for
developers, who want to  assume that library code behaves in a  predictable fashion; nondeterminism is
vexing in debugging; nondeterminism is \emph{often disastrous} for large-scale 
automated testing.

@@

Complex
modern software systems usually include a large set of
\emph{regression tests}.  A regression test suite is a
set of tests that can be run every time code
is modified, to ensure that the modification has not broken the system.  \emph{Flaky tests} \cite{miccoflaky} are regression
tests that fail in an intermittent, unreliable fashion.  The essence of a flaky
test is that, for the same snapshot of test code and code under test, it sometimes fails and sometimes passes: the pass/fail result (disposition) of the
test is not a deterministic property of the test code, code under
test, and testing environment.  This produces three serious
problems: first, a flaky test often wastes developer time and delays
software changes by forcing the investigation of \emph{correct} code-under-test.  Second, failures in flaky tests are often ignored, and therefore serious software faults missed.  Finally, to mitigate the problem, flaky tests are often
run multiple times, wasting computing resources and
delaying code changes.  Flaky tests are, for us, simply a special case the 
\emph{horizontal} nondeterminism 
described below.  Our focus is on detecting sources of 
flaky-ness, \emph{before} they propagate.
\begin{comment}
  As an analogy, we note that  a canary in a coal mine is of
little use if canaries frequently become ill for reasons unrelated to
the presence of toxic gases.  Mining may stop for no good reason, or
miners may learn to ignore the canary, leading to tragedy; a third,
more ``practical'' option is that miners may carry so many redundant
canaries into the coal mine that canary-care itself becomes a serious burden.
\end{comment}

### 2020-07-15 ISSTA/relatedwork.tex
shrink
down to size

At a broad level, the topic of uncertainty (and thus nondeterminism) in software engineering has been addressed by Garlan \cite{GarlanUncertain}, Elbaum and Rosenblum \cite{Unknowns,lu2015roundtable}, and Ziv and Richardson \cite{UncertaintyPrinciple}.  There is a general agreement that as systems become more complex, more distributed, and more \emph{statistical}, these problems will only grow \cite{lu2015roundtable}.
Gao et al. examined the question of how to make tests repeatable \cite{Gao:2015:MSU:2818754.2818764}, in the context of system-wide user interaction tests, focusing on systemic factors in contrast to our focus on  the library level.  
Shi et. al \cite{DetermImp} examined what might be seen as the opposite, problem:  they detect code that assumes a deterministic implementation of a non-deterministic specification.  %E.g., they detect instances when  iteration through items of a set is explicitly not guaranteed to provide any particular order of items, but code depends on the order. Determinism is also sometimes used as a property in domain-specific testing, e.g. for shader compilers \cite{shader}.

The problem of test nondeterminism is closely related to the issue of flaky tests \cite{miccoflaky, luo2014empirical,palomba2017does,listfieldtestanalysis}.  How to handle flaky tests in practice (when they cannot be avoided) is a major issue in Google-scale continuous testing \cite{memon2017taming}, and, as Memon et al. describe, the problem of flaky tests influences the general design of Google test automation.
Previous work on flaky tests has either focused on test inter-dependence \cite{LamZE2015}, or large-scale empirical examination \cite{luo2014empirical,palomba2017does}.  
Bell et al. proposed DeFlaker \cite{bell2018d}, which makes flaky tests much easier to detect by relying on the observation that if a test fails, and does not cover any changed code then the test is likely flaky.   iDFlakies \cite{idflakies} is a framework and dataset for flaky tests, but again focuses on whole-tests, and detecting flakiness by actually observing it.  Their work shows the order-dependence accounts for a little more than 50\% of flaky test causes, and they classify the rest as due to ``concurrency, timeouts,
network/IO, etc.''---the causes that our approach focuses on identifying.

We introduce the first variation of
delta-debugging that properly handles probabilistic reduction criteria, in line with Harman and O'Hearn's proposal to simply accept that ``All Tests Are Flaky'' \cite{StartupstoScaleups}, and work with probabilistically failing tests..
Vertical/failure determinism is also a less-studied concept, and to our knowledge our formulation is novel. The kinds of errors that are exposed, however, are not new.  In particular, we believe that many such faults are related to the propagation of error conditions, which has been studied using static analysis \cite{FileProp}.

@@

\begin{comment}
Finally, Cotroneo et al. \cite{CompBugs,FaultTriggers}, and Grottke and Trivedi \cite{GrottkeBugs} have followed on early work on understanding bugs and how they manifest, including transient \cite{Transient} software faults.  This work informs our attempt to identify sources of nondeterminism, and should provide other, context and project-specific, sources that could be introduced into our general framework.
\end{comment}

>>>>

The general topic of nondeterminism in software engineering has long been considered important  \cite{Unknowns,UncertaintyPrinciple}. 
Gao et al. examined the question of how to make tests repeatable \cite{Gao:2015:MSU:2818754.2818764}, in the context of system-wide user interaction tests, focusing on systemic factors.  
Shi et. al \cite{DetermImp} examined code that wrongly enforces a deterministic implementation.  %E.g., they detect instances when  iteration through items of a set is explicitly not guaranteed to provide any particular order of items, but code depends on the order. Determinism is also sometimes used as a property in domain-specific testing, e.g. for shader compilers \cite{shader}.

The problem of test nondeterminism is closely related to the issue of flaky tests \cite{miccoflaky, luo2014empirical,palomba2017does,listfieldtestanalysis}.  How to handle flaky tests in practice is a major issue in Google-scale continuous testing \cite{memon2017taming}.
Previous work on flaky tests has either focused on test inter-dependence \cite{LamZE2015}, or large-scale empirical examination \cite{luo2014empirical,palomba2017does}.  
Bell et al. proposed DeFlaker \cite{bell2018d}, which makes flaky tests much easier to detect by relying on the observation that if a test fails, and does not cover any changed code then the test is likely flaky.   iDFlakies \cite{idflakies} is a framework and dataset for flaky tests, but again focuses on whole-tests, and detecting flakiness by actually observing it.

We introduce the first variation of
delta-debugging that properly handles probabilistic reduction criteria, in line with Harman and O'Hearn's proposal to simply accept that ``All Tests Are Flaky'' \cite{StartupstoScaleups}, and work with probabilistically failing tests..
Vertical/failure determinism is also a less-studied concept, and to our knowledge our formulation is novel. The kinds of errors that are exposed, however, are not new, e.g., faults related to the propagation of error conditions \cite{FileProp}.

@@
